Stream Members Only:\\ Data-Driven Characterization of Stellar Streams with Mixture Density Networks

### 2024-04-21 src/tex/ms.tex
v2 - cleanup
Signed-off-by: nstarman <nstarman@users.noreply.github.com>

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath, amssymb}
\usepackage{gensymb}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage[inline]{enumitem}
\usepackage{cleveref}
\usepackage{transparent}
\usepackage{todonotes}
\usepackage{array}
\usepackage{float}

@@

\newcommand{\galaxy}[1]{
    \ifthenelse{\equal{#1}{}}{
        \includegraphics[width=9pt]{static/galaxy.png}
    }{
        \includegraphics[width={#1}pt]{static/galaxy.png}
    }%
}

@@

\newcommand{\TODO}[1]{{\textcolor{red}{#1}}}
\newcommand{\JN}[1]{\TODO{#1}}

@@

% Title
\title{Does this Star Belong? Data-Driven Characterization of Stellar Streams \\ with Mixture Density Networks}

% Author list
\author[0000-0003-3954-3291]{Nathaniel Starkman}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto}

\author[0000-0001-8042-5794]{Jacob Nibauer}
\altaffiliation{Co-first author.}
\affiliation{Department of Astrophysical Sciences, Princeton University}

\author[0000-0001-6855-442X]{Jo Bovy}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto}

\author[0000-0003-3613-0854]{Jeremy J. Webb}
\affiliation{Division of Natural Science, Department of Science, Technology and Society, York University}

% \author[0000-0002-7846-9787]{Ana Bonaca}
% \affiliation{Carnegie Observatories: Pasadena, California, US}

@@

% Abstract with filler text
\begin{abstract}
    Stellar streams are sensitive probes of the Milky Way's gravitational
    potential. The mean track of a stream constrains global properties of the
    potential, while its fine-grained surface density contains information about
    galactic substructure. A precise characterization of streams from potentially noisy data marks a crucial step in constraining galactic structure across orders of magnitude in mass scales. Here we present a new method for constructing a smooth probability
    density model of stellar streams using all of the available astrometric and
    photometric data. To characterize the stream's morphology and kinematics, we
    utilize mixture density networks to construct a flexible representation of
    the stream and its on-sky track, width, stellar number density, and
    kinematic distribution. In photometric coordinates we model each stream as a
    single-stellar population, with a distance track that is simultaneously
    estimated from the stream's inferred distance modulus (using photometry) and
    parallax distribution (using astrometry). Normalizing flows are used to characterize the distribution of background stars. We apply the method to the stream
    \stream{GD-1}, and the tidal tails of \stream{Palomar 5}. For both streams we obtain a catalog
    of stellar membership probabilities that are made publicly available.
    Importantly, our model is capable of handling data with incomplete
    phase-space observations, making our method applicable to the growing census
    of Milky Way stellar streams. When applied to a population of streams, the
    resulting membership probabilities from our model form the required input to
    infer the Milky Way's dark matter distribution from the scale of the stellar
    halo down to subhalos. 
\end{abstract}

@@

    Stellar streams are the disrupted remnants of globular clusters and
    satellite galaxies, and their census has grown extensively in the Milky Way
    thanks to all-sky astrometric missions like \Gaia{}
    \citep{GaiaCollaboration+2016, GaiaCollaboration+2023} and low-surface
    brightness photometric surveys (e.g., \citealt{Ibata+1998,
    GrillmairDionatos2006, Shipp+2018, Ibata+2021}).  The formation of streams
    can be traced back to a progenitor (a globular cluster or satellite galaxy)
    whose stars are gradually lost to the tidal field of the host galaxy. The
    stars extend along a series of similar orbits, called tidal tails, which can
    sometimes span several tens of degrees across the sky. The orbital proximity
    of stars belonging to a stream makes them sensitive tracers of the mass
    distribution of galaxies, sourced by both the baryonic and dark matter
    components.

    The average properties of a stream (e.g., the mean position and velocity
    track of the tidal tails) can be used to reconstruct the global dark matter
    distribution of the stellar halo (e.g., \citealt{Johnston+1999, Bonaca+2014,
    Bovy2014, Bovy+2016, Nibauer+2022, Koposov+2023}), while local variations in
    the stream track and surface density encode information about the stream's
    encounters with baryonic structures such as the stellar bar (e.g.,
    \citealt{Erkal+2017, Pearson+2017}), and non-luminous structures like dark
    matter subhalos (e.g., \citealt{Ibata+2002, Johnston+2002,
    Carlberg+2012-74820C, Bovy+2017, Bonaca+2019, Hermans+2021}).

    While several methods have been developed to model stream properties as a
    function of global halo characteristics (e.g., the mass distribution,
    flattening; \citealt{Johnston+1999,2008MNRAS.386L..47B,Koposov+2010,2013MNRAS.433.1826S,Bovy2014,Nibauer+2022}) and local substructures (e.g., the bar's pattern
    speed and the mass of subhalo perturbers; \citealt{2014A&A...563A..60A,Pearson+2017,2016MNRAS.460..497H,Bovy+2017,Bonaca+2019}), these methods often
    rely on the existence of a homogenized stream-tracer population that
    characterizes a given stream across each of the observable phase-space
    dimensions. Additionally, because even the fine-grained properties of
    streams are influenced by the detailed mass distribution of its host galaxy,
    it is important to propagate errors both in measurement uncertainty and
    membership uncertainty (i.e., which stars belong to the stream versus the
    background) when attempting to model a stream. Otherwise, a poorly
    characterized stream could bias constraints on, e.g., the total mass of the
    galaxy, the dark matter halo shape, and the mass of subhalo perturbers. In
    order to generate more precise constraints on the structure of the galaxy,
    it is therefore imperative that a statistically sound and homogenized
    catalog of stellar streams is produced.

@@

    There are a few previous works which have modeled stellar streams in this
    context. \citet{Patrick+2022} developed the method introduced by
    \citet{Erkal+2017}, which is a data-driven, spline-based method for modeling
    the photometric properties and on-sky positions of previously identified
    streams. Their method works by fitting the stream's stellar surface density
    with a series of splines, capturing both surface brightness variations along
    the stream and changes in the position of the stream track and width. Their
    method also utilizes an isochrone fit to a given stream, from which a
    distance modulus can be derived. While this method is very successful at
    characterizing the photometric properties of a stream, it does not consider
    the kinematic dimensions which have become increasingly well-measured with
    successive data releases from \Gaia{}, as well as targeted observations.
    From a population of RR Lyrae stars, \citet{Price-Whelan+2019} utilized a
    mixture model to fit the globular cluster stream Pal-5 in the space of
    proper motions and heliocentric distances. The mixture modeling approach
    enables a simultaneous fit to both the stream and background density,
    allowing for better separation of the stream from the background
    distribution. However, the track of the stream in on-sky coordinates is not
    fit. The method identified 27 RR Lyrae stars consistent with being members
    of the stream. While useful for determining the average properties of a
    stream (e.g., its mean distance track), a limited population of RR Lyrae
    stars are not sufficient to capture the small-scale properties of streams
    expected to inform constraints on, e.g., a population of perturbers in the
    galaxy.

@@

        The paper is organized as follows.  In \autoref{sec:method} we introduce
        our method for characterizing stream tracks and computing stellar
        membership.  We discuss initial data processing in
        \autoref{sub:method:framing_the_data} and build a probabilistic
        framework to model the stream and background in
        \autoref{sub:method:likelihood_setup}-\autoref{sub:methods:priors}.  In
        \autoref{sec:results_mock} we apply our method to simulated data to
        demonstrate the model with a known ground truth dataset, then to the
        stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and
        \stream{Pal\,5} (\autoref{sec:results_pal5}).  We compare our approach
        to characterizing stellar streams to other methods in
        \autoref{sec:discussion}.  We conclude in \autoref{sec:conclusions},
        discussing the results of this work and future directions.

@@

\section{Method} \label{sec:method}

    To characterize stellar stream tracks and for computing membership
    probabilities from kinematic and photometric measurements of stars, we
    develop a flexible Bayesian modeling framework that incorporates all of the
    available kinematic information for a field of stars, including those with
    only partial phase-space measurements. Below, we offer a concise overview of
    our approach, followed by a more in-depth technical explanation in
    subsequent sections.  We start, in \autoref{sub:method:framing_the_data}, by
    transforming the astrometric data into a stream-oriented on-sky reference
    frame ($\phi_1, \phi_2, \parallax, \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for
    each stream. In this frame, the stream track lies along $\phi_1$, and the
    stream density may be described by distributions in the other parameters as
    injective functions of $\phi_1$.  Likewise, the background of the Galactic
    field may be described with $\phi_1$-conditioned distributions.  We will
    also model the stream in photometric coordinates. Additional features, e.g.
    metallicity, can also be modeled along the stream and background as
    functions of $\phi_1$, though we leave this to a future work.

    The distributions for the stream and background are collected in a single
    mixture model that describes the entire field.  We take the parameters of
    the mixture model -- the mixture weights and component distribution
    parameters -- to be general (continuous) functions of the conditioning
    variable, $\phi_1$. This is done using a feed-forward neural network that
    takes the conditioning variable as its input and outputs the parameter
    value(s) that characterize the distribution of stars at a $\phi_1$ ``slice".
    Mixture models of conditional probability distributions using neural
    networks are called Mixture Density Networks (MDNs) \citep{Bishop+1994}.
    Since the model coefficients may evolve with $\phi_1$, MNDs are well suited
    to characterize streams (and the background) with all their variations in
    position, width, and linear density.

@@

        While it would be ideal to characterize streams in this Galactocentirc
        frame, transformation to the frame requires knowledge of the distance to
        the Galactic center  \citep{GRAVITY2018, BennettBovy2019, Leung+2022},
        the solar motion around the Galactic center, and accurate distances to
        field stars (among other uncertain quantities). In a heliocentric frame,
        e.g. the standard ICRS\footnote{International Celestial Reference
        System} \citep{ICRS1997}, projection effects can obscure the shape of a
        stream leading to a non-trivial stream track. Therefore in a
        heliocentric frame, there is no assurance of effecting a rotation like
        in the Galactocentric frame that would similarly align the stream in an
        observer's coordinates such that $\phi_2(\phi_1) \approx 0$.
        Thankfully, most streams are adequately distant from both the Galactic
        center and the observer, or exhibit orientations such that a rotation a)
        can be established or b) holds valid. For most of the best-studied
        globular cluster streams -- \stream{GD-1}, \stream{Palomar 5},
        \stream{Jhelum}, etc. -- this holds true. However for some streams of
        dwarf galaxies, like the Magellanic stream which wraps multiple times
        around the Milky Way \citep{WannierWrixon1972}, no such transformation
        is possible. More generally, a stream's ``sky"-projected path may be
        kinked, e.g. by subhalo interactions, such that it is a many-to-one
        function in $\phi_1$.

        In this work we treat with the vast majority of streams, which have a
        ``sky" transformation such that the stream is an injective function in
        $\phi_2(\phi_1)$.  Since we are examining known streams this
        transformation is known \textit{a priori}, or may be constructed during
        pre-processing \citep[see][\S2.1]{Starkman+2023}.  We note that, with
        sufficient prior knowledge, many-to-one functions may be broken into
        injective segments and their connection constrained by a prior. Thus the
        methods developed subsequently may be extended and applied even to
        wrapped and kinked streams. However, we leave extending the model
        framework to a future work.

@@

        \begin{figure*}
            \script{pgm.py}
            \centering
            \includegraphics[scale=0.6]{figures/pgm.pdf}
            \caption{%
                Probabilistic Graphical Models (PGMs) of the Mixture Density
                Networks used in this work.
                %
                \textbf{Left}: A PGM of the stream model. The full dataset
                $x_n^{obs}$ is made of the observed astrometry $w_n^{obs}$ and
                photometry $m_n^{obs}$, along with the corresponding
                observational errors. We model the data as a mixture of
                Gaussians (in astrometry + distance modulus), with the Gaussians
                indexed by $q$. The weight $\alpha$ determines the weighted
                contribution of each Gaussian to each datum. All models are
                conditioned on $\phi_1$.
                %
                \textbf{Right}: a PGM of the background model. The data is
                identical to that of the stream, however the model is not a
                mixture of Gaussians, but of a variety of distributions. We
                further distinguish between trainable distribution parameters
                (circles) and fixed distributions (points), like a pre-trained
                normalizing flow background model, explained in further detail
                in \autoref{sub:method:pre-training_distributions}. Alike to the
                stream model, all background models are conditioned on $\phi_1$.
                % 
            }
            \label{fig:PGM}
        \end{figure*}

        Mixture models (MMs) are a statistical method to represent a population
        as a set of a sub-populations specified by probability distributions.
        These distributions are characterized by their parameters and the
        mixture by a corresponding set of mixture coefficients. Scalar mixture
        models may be extended to model completely general conditional
        distributions by making the distribution parameters and mixture
        coefficients into functions of a conditional input
        \citep{McLachlanBasford1989}.  Mixture Density Networks (MDNs) are a
        machine learning method to implement these general MM, taking the
        parameters and coefficients to be the outputs of feed-forward neural
        networks and as such, general continuous functions \citep{Bishop1994}.
        Functional parameters allow the model to capture variation over those
        parameters that is not possible with scalar-valued models. Moreover, by
        using neural networks the inferred functional forms are ``model-free"
        and driven entirely by the data.

        For the MDN we take $x_{\cdot, 0}$, which is the first feature column of
        the data $\mbf{X}$, as the `independent' coordinate over which the MDN
        parameters are conditioned. In the context of streams $x_{\cdot, 0}
        \equiv \phi_1$ is the longitude coordinate in a reference frame rotated
        such that the stream is aligned along $\phi_1$. The parameters of the
        stream, the background field, and the mixing coefficients of the two are
        now general data-driven continuous functions along $\phi_1$.

        The PDF of a general MDN is
        \begin{equation} \label{eq:general_mixture_network}
            \!\!\! \pdf(\nth{\mbs{x}} | \mbs{\theta}(\phi_1))
            \!=\! \sum_{\mrm{q} \in \mbs{I_Q}} \! \qth{f}(\phi_1) \cmp{q}{\pdf}(\nth{\mbs{x}}|\qth{\mbs{\theta}}(\phi_1)),
            % \prior(\nth{\mbs{x}}, \phi_1)
        \end{equation}
        where $\cmp{q}{\pdf}$ is the PDF of the $q$-th model and $f_q$ its
        mixing coefficient for all $q$ models in the set of models indexed by
        $\mbs{I_Q}$. $\mbs{\theta}$ is the $\phi_1$-conditional output of the
        feed-forward neural network, and $\nth{\mbs{x}}$ the data on the $n$-th
        star in $\mbs{X}$, not including $\phi_1$. For convenience we drop the
        implied subscript $n$ in $\phi_1$, the explicit notation of
        $\mbs{\theta}$'s dependence on $\phi_1$, and the index set $\mbs{I}_X$
        in summations over its elements $x$.

        The mixture coefficients $\qth{f}$ are normalized such that
        \begin{equation} \label{eq:mixture_weight_sum}
            \sum_{\mrm{q}} \qth{f}(\phi_1) = 1, \qquad \forall \phi_1
        \end{equation}
        In practice this is enforced by defining a background
        \begin{equation} \label{eq:mixture_weight_normalization}
            f_{b}(\phi_1) = 1 - \sum_{\rm{q}\in \mbs{I_Q}\backslash \{b\}} \qth{f}(\phi_1),
        \end{equation}
        the summation over all models except for some ``background" index $b \in
        \mbs{I_Q}$.  In the context of Galactic streams the ``foreground" is the
        stream itself, while the background is the Galaxy - bulge, bar, disc,
        halo - and additional structures, like globular clusters and dwarf
        galaxies, that are not associated with the stream itself.  We discuss
        the mixture weights further in \autoref{sub:method:mixture_weight}.

        In this work we consider primarily astrometric and photometric data,
        (notated $\mbs{W}$ and $\mbs{M}$, respectively), but note that the MDN
        framework may be extended to arbitrary feature dimensions. Assuming the
        conditional independence of the astrometry and photometry\footnote{ For
        \Gaia{} data this is true in practice but not in detail: there exist
        small correlations between the astrometric and photometric data.}, we
        split each model into linearly separable astrometric and photometric
        models $\cmp{q}{\pdf} = \cmp{q,w}{\pdf} \cmp{q,m}{\pdf}$.

@@

    %\newpage
    \subsection{Missing Phase-Space Observations}
    \label{sub:method:missing_data}

        Streams have a diversity of measurement coverage: several have full 6-D
        astrometrics as well as photometry and chemical abundances available
        \citep[e.g.,][]{Koposov+2019, Antoja+2020, Li+2022}, while most have
        partial coverage, lacking some feature dimensions.  Most streams in
        \Gaia \citep{Gaia2016, Gaia2023}, for instance, have proper motions but
        not radial velocities nor high signal-to-noise parallax measurements.
        Follow-up surveys, like \dataarchive{S5} \citep{Li+2019}, add many
        feature dimensions, however only for a select set of streams. Given the
        diversity of coverage it is important that the likelihood model --
        \autoref{eq:membership_prob} -- account for missing data.

>>>>

\documentclass[twocolumn, linenumbers]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath, amssymb}
\usepackage{gensymb}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage[inline]{enumitem}
\usepackage{cleveref}
\usepackage{transparent}
\usepackage{array}
\usepackage{float}

@@

@@

@@

% Title
\title{Stream Members Only:\\ Data-Driven Characterization of Stellar Streams with Mixture Density Networks}

% Author list
\author[0000-0003-3954-3291]{Nathaniel Starkman}
\altaffiliation{Co-first author.}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto, 50 St. George St, Toronto ON M5S 3H4, Canada}

\author[0000-0001-8042-5794]{Jacob Nibauer}
\altaffiliation{Co-first author.}
\affiliation{Department of Astrophysical Sciences, Princeton University, 4 Ivy Ln, Princeton, NJ 08544, USA}

\author[0000-0001-6855-442X]{Jo Bovy}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto, 50 St. George St, Toronto ON M5S 3H4, Canada}

\author[0000-0003-3613-0854]{Jeremy J. Webb}
\affiliation{Department of Science, Technology and Society, York University, 4700 Keele St, Toronto ON M3J 1P3, Canada}

\author[0000-0001-6584-6144]{Kiyan Tavangar}
\affiliation{Department of Astronomy, Columbia University, 538 West 120th Street, New York, NY 10027, USA}

\author[0000-0003-0872-7098]{Adrian Price-Whelan}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

\author[0000-0002-7846-9787]{Ana Bonaca}
\affiliation{The Observatories of the Carnegie Institution for Science, 813 Santa Barbara Street, Pasadena, CA 91101, USA}

@@

\begin{abstract}

    Stellar streams are sensitive probes of the Milky Way's gravitational
    potential. The mean track of a stream constrains global properties of the
    potential, while its fine-grained surface density constrains galactic
    substructure. A precise characterization of streams from potentially noisy
    data marks a crucial step in inferring galactic structure, including the
    dark matter, across orders of magnitude in mass scales. Here we present a
    new method for constructing a smooth probability density model of stellar
    streams using all of the available astrometric and photometric data. To
    characterize a stream's morphology and kinematics, we utilize mixture
    density networks to represent its on-sky track, width, stellar number
    density, and kinematic distribution. We model the photometry for each stream
    as a single-stellar population, with a distance track that is simultaneously
    estimated from the stream's inferred distance modulus (using photometry) and
    parallax distribution (using astrometry). We use normalizing flows to
    characterize the distribution of background stars. We apply the method to
    the stream \stream{GD-1}, and the tidal tails of \stream{Palomar 5}. For
    both streams we obtain a catalog of stellar membership probabilities that
    are made publicly available.  Importantly, our model is capable of handling
    data with incomplete phase-space observations, making our method applicable
    to the growing census of Milky Way stellar streams.

\end{abstract}

@@

    Stellar streams are the disrupted remnants of globular clusters and
    satellite galaxies, and their census has grown extensively in the Milky Way
    thanks to all-sky astrometric missions like \Gaia{}
    \citep{GaiaCollaboration+2016, GaiaCollaboration+2023} and low-surface
    brightness photometric surveys (e.g., \citealt{Ibata+1998,
    GrillmairDionatos2006, GrillmairCarlin2016, Shipp+2018, Ibata+2021}).  The formation of streams
    can be traced back to a progenitor (a globular cluster or satellite galaxy)
    whose stars are gradually lost to the tidal field of the host galaxy. The
    stars extend along a series of similar orbits, called tidal tails, which can
    sometimes span several tens of degrees across the sky. The orbital proximity
    of stars belonging to a stream makes them sensitive tracers of the mass
    distribution of galaxies, sourced by both the baryonic and dark matter
    components.

    The average properties of a stream along its track (e.g., the mean position and velocity
    track of the tidal tails) can be used to reconstruct the global dark matter
    distribution of the stellar halo (e.g., \citealt{Johnston+1999, Bonaca+2014,
    Bovy2014, Bovy+2016, Nibauer+2022, Koposov+2023}), while local variations in
    the stream track and surface density encode information about the stream's
    progenitor \citep{Kupper+2012, Sanders+2013, Price-Whelan+2014}, encounters
    with baryonic structures such as the stellar bar (e.g., \citealt{Erkal+2017,
    Pearson+2017}), and non-luminous structures like dark matter subhalos (e.g.,
    \citealt{Ibata+2002, Johnston+2002, Carlberg+2012-74820C, Bovy+2017,
    Bonaca+2019, Hermans+2021}).

    While several methods have been developed to model stream properties as a
    function of global halo characteristics (e.g., the mass distribution,
    flattening; \citealt{Johnston+1999, Binney2008, Koposov+2010,
    SandersBinney2013, Bovy2014, Nibauer+2022}) and local substructures (e.g.,
    the bar's pattern speed and the mass of subhalo perturbers;
    \citealt{Antoja+2014, Hattori+2016, Price-Whelan+2016, Bovy+2017,
    Pearson+2017, Bonaca+2019, Malhan+2021}), these methods often rely on the existence of a
    homogenized stream-tracer population that characterizes a given stream
    across each of the observable phase-space dimensions.  Additionally, because
    even the fine-grained properties of streams are influenced by the detailed
    mass distribution of its host galaxy, it is important to propagate errors
    both in measurement uncertainty and membership uncertainty (i.e., which
    stars belong to the stream versus the background) when attempting to model a
    stream. Otherwise, a poorly characterized stream could bias constraints on,
    e.g., the total mass of the galaxy, the dark matter halo shape, and the mass
    of subhalo perturbers.  In order to generate more precise constraints on the
    structure of the galaxy, it is therefore imperative that a statistically
    sound and homogenized catalog of stellar streams is produced.

@@

    There are a few previous works which have modeled stellar streams in this
    context. \citet{Patrick+2022} developed the method introduced by
    \citet{Erkal+2017}, which is a data-driven, spline-based method for modeling
    the photometric properties and on-sky positions of previously identified
    streams. Their method works by fitting the stream's stellar surface density
    with a series of splines, capturing both surface brightness variations along
    the stream and changes in the position of the stream track and width. Their
    method also utilizes an isochrone fit to a given stream, from which a
    distance modulus can be derived. While this method is very successful at
    characterizing the photometric properties of a stream, it does not consider
    the kinematic dimensions which have become increasingly well-measured with
    successive data releases from \Gaia{}, as well as targeted observations.
    From a population of RR Lyrae stars, \citet{Price-Whelan+2019} utilized a
    mixture model to fit the globular cluster stream Pal-5 in the space of
    proper motions and heliocentric distances.  The mixture modeling approach
    enables a simultaneous fit to both the stream and background density,
    allowing for a cleaner selection of stream stars. However, the track of the
    stream in on-sky coordinates is not fit. The method identified 27 RR Lyrae
    stars consistent with being members of the stream. While useful for
    determining the average properties of a stream (e.g., its mean distance
    track), a limited population of RR Lyrae stars are not sufficient to capture
    the small-scale properties of streams expected to inform constraints on,
    e.g., a population of perturbers in the galaxy.

@@

        The paper is organized as follows.  In \autoref{sec:method} we introduce
        our method for characterizing stream tracks and computing stellar
        membership probabilities.  We discuss initial data processing in
        \autoref{sub:method:framing_the_data} and build a probabilistic
        framework to model the stream and background in
        \autoref{sub:method:likelihood_setup}-\autoref{sub:methods:priors}.  In
        \autoref{sec:results_mock} we apply our method to simulated data to
        demonstrate the model with a known ground truth dataset, then to the
        stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and
        \stream{Pal\,5} (\autoref{sec:results_pal5}).  We compare our model to
        other methods in \autoref{sec:discussion}.  We conclude in
        \autoref{sec:conclusions}, discussing the results of this work and
        future directions.

@@

\newpage
\section{Method} \label{sec:method}

    Below, we offer a concise overview of our approach, followed by a more
    in-depth technical explanation in subsequent sections.  We start, in
    \autoref{sub:method:framing_the_data}, by transforming the astrometric data
    into a stream-oriented on-sky reference frame ($\phi_1, \phi_2, \parallax,
    \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for each stream.  In this frame, the
    stream track lies along $\phi_1$, and the stream density may be described by
    distributions in the other parameters as injective functions of $\phi_1$
    (called conditional distributions).  Likewise, the background of the
    Galactic field may be described with $\phi_1$-conditioned distributions.  We
    will also model the stream in photometric coordinates.  Additional features,
    e.g.  metallicity, can also be modeled along the stream and background as
    functions of $\phi_1$, though we leave this to a future work.

    The distributions for the stream and background are collected in a single
    mixture model that describes the entire field.  We take the parameters of
    the mixture model --- the mixture weights and component distribution
    parameters --- to be general (continuous) functions of $\phi_1$, the
    conditioning variable. This is done using a feed-forward neural network that
    takes $\phi_1$, the conditioning variable, as its input and outputs the
    parameter value(s) that characterize the distribution of stars at a $\phi_1$
    ``slice".  Mixture models of conditional probability distributions using
    neural networks are called Mixture Density Networks (MDNs)
    \citep{Bishop+1994}.  Since the model coefficients may evolve with $\phi_1$,
    MNDs are well suited to characterize streams (and the background) with all
    their variations in position, width, and linear density.

@@

        While it would be ideal to characterize streams in this Galactocentric
        frame, transformation to the frame requires knowledge of the distance to
        the Galactic center  \citep{GRAVITY2018, BennettBovy2019, Leung+2022},
        the solar motion around the Galactic center, and accurate distances to
        field stars (among other uncertain quantities). In a heliocentric frame,
        e.g. the standard ICRS\footnote{International Celestial Reference
        System} \citep{ICRS1997}, projection effects can obscure the shape of a
        stream leading to a non-trivial stream track. Therefore in a
        heliocentric frame, there is no assurance of effecting a rotation like
        in the Galactocentric frame that would similarly align the stream in an
        observer's coordinates such that $\phi_2(\phi_1) \approx 0$.
        Thankfully, most streams are adequately distant from both the Galactic
        center and the observer, or exhibit orientations such that a rotation a)
        can be established or b) holds valid. For most of the best-studied
        globular cluster streams -- \stream{GD-1}, \stream{Palomar 5},
        \stream{Jhelum}, etc. -- this holds true. However for some streams of
        dwarf galaxies, like the Magellanic and Sagittarius streams which each
        wrap around the Milky Way \citep{Newberg+2002, Majewski+2003,
        WannierWrixon1972}, no such transformation is possible. More generally,
        a stream's ``sky"-projected path may be kinked, e.g. by subhalo
        interactions, such that it is a many-to-one function in $\phi_1$.

        In this work we consider the vast majority of streams that can be
        described as an injective function, $\phi_2(\phi_1)$.  Since we are
        examining known streams this transformation is known \textit{a priori},
        or may be constructed during pre-processing
        \citep[see][\S2.1]{Starkman+2023}.  We note that, with sufficient prior
        knowledge, many-to-one functions may be broken into injective segments
        and their connection constrained by a prior. Thus the methods developed
        here may be extended and applied even to wrapped and kinked streams.
        However, we leave extending the model framework to a future work.

@@

        \begin{figure*}
            \script{pgm.py}
            \centering
            \includegraphics[scale=0.6]{figures/pgm.pdf}
            \caption{%
                Probabilistic Graphical Models (PGMs) of the Mixture Density
                Networks used in this work.
                %
                \textbf{Left}: A PGM of the stream model. The full dataset
                $x_n^{obs}$ is the aggregation of the observed astrometry
                $w_n^{obs}$ and photometry $m_n^{obs}$, along with the
                corresponding observational errors. We model the data as a
                mixture of Gaussians (in astrometry + distance modulus), with
                the Gaussians indexed by $q$. The weight $\alpha$ determines the
                weighted contribution of each Gaussian to each datum. All models
                are conditioned on $\phi_1$.
                %
                \textbf{Right}: a PGM of the background model. The data are
                identical to that of the stream, however the model is not a
                mixture of Gaussians, but of a variety of distributions. We
                further distinguish between trainable distribution parameters
                (circles) and fixed distributions (points), like a pre-trained
                normalizing flow background model, explained in further detail
                in \autoref{sub:method:pre-training_distributions}. Alike to the
                stream model, all background models are conditioned on $\phi_1$.
                % 
            }
            \label{fig:PGM}
        \end{figure*}

        Mixture models are a statistical method to represent a population as a
        set of a sub-populations specified by probability distributions.  These
        distributions are characterized by their parameters and the mixture by a
        corresponding set of mixture coefficients. Scalar mixture models may be
        extended to model completely general conditional distributions by making
        the distribution parameters and mixture coefficients into functions of a
        conditional input \citep{McLachlanBasford1989}.  Mixture Density
        Networks (MDNs) are a machine learning method to implement these general
        mixture models, taking the parameters and coefficients to be the outputs
        of feed-forward neural networks and as such, general continuous
        functions \citep{Bishop1994}.  Functional parameters allow the model to
        capture variation over those parameters that is not possible with
        scalar-valued models. Moreover, by using neural networks the inferred
        functional forms are ``model-free" and driven entirely by the data.

        For the MDN we take $x_{\cdot, 0}$, the first feature column of the data
        $\{\mbf{x}\}$, as the `independent' coordinate over which the MDN
        parameters are conditioned. In the context of streams $x_{\cdot, 0}
        \equiv \phi_1$ is the longitude coordinate in a reference frame rotated
        such that the stream is aligned along $\phi_1$. The parameters of the
        stream, the background field, and the mixing coefficients of the two are
        now general data-driven continuous functions along $\phi_1$.

        The PDF of a general MDN is
        \begin{equation} \label{eq:general_mixture_network}
            \!\!\! \pdf(\nth{\mbs{x}} | \mbs{\theta}(\phi_1))
            \!=\! \sum_{\mrm{q} \in \mbs{I_Q}} \! \qth{f}(\phi_1) \cmp{q}{\pdf}\left(\nth{\mbs{x}}|\qth{\mbs{\theta}}(\phi_1)\right),
        \end{equation}
        where $\cmp{q}{\pdf}$ is the PDF of the $q$-th model and $f_q$ its
        mixing coefficient for all $q$ models in the set of models indexed by
        $\mbs{I_Q}$. $\mbs{\theta}$ is the $\phi_1$-conditional output of the
        feed-forward neural network, and $\nth{\mbs{x}}$ the data on the $n$-th
        star in $\{\mbs{x}\}$, not including $\phi_1$. For convenience we drop
        the implied subscript $n$ in $\phi_1$, the explicit notation of
        $\mbs{\theta}$'s dependence on $\phi_1$, and the index set $\mbs{I}_X$
        in summations over its elements $x$.

        The mixture coefficients $\qth{f}$ are normalized such that
        \begin{equation} \label{eq:mixture_weight_sum}
            \sum_{\mrm{q}} \qth{f}(\phi_1) = 1, \qquad \forall \phi_1
        \end{equation}
        In practice this is enforced by defining a background
        \begin{equation} \label{eq:mixture_weight_normalization}
            f_{b}(\phi_1) = 1 - \sum_{\rm{q}\in \mbs{I_Q}\backslash \{b\}} \qth{f}(\phi_1),
        \end{equation}
        the summation over all models except for some ``background" index $b \in
        \mbs{I_Q}$.  In the context of Galactic streams the ``foreground" is the
        stream itself, while the background is the Galaxy --- bulge, bar, disc,
        halo --- and additional structures, like globular clusters and dwarf
        galaxies, that are not associated with the stream itself.  We discuss
        the mixture weights further in \autoref{sub:method:mixture_weight}.

        In this work we consider primarily astrometric and photometric data,
        (notated $\{\mbs{w}\}$ and $\{\mbs{m}\}$, respectively), but note that
        the MDN framework may be extended to arbitrary feature dimensions.
        Assuming the conditional independence of the astrometry and
        photometry\footnote{ For \Gaia{} data this is true in practice but not
        in detail: there exist small correlations between the astrometric and
        photometric data \citep[see e.g. the \href{https://www.cosmos.esa.int/web/gaia/science-performance\#astrometric\%20performance}{\Gaia{} website} or][]{GaiaDR2Astrometry2018}.}, we split each model into linearly separable
        astrometric and photometric models $\cmp{q}{\pdf} = \cmp{q,w}{\pdf}
        \cmp{q,m}{\pdf}$.

@@

    \subsection{Missing Phase-Space Observations}
    \label{sub:method:missing_data}

        Streams have a diversity of measurement coverage: several have full 6-D
        astrometrics as well as photometry and chemical abundances available
        \citep[e.g.,][]{Koposov+2019, Antoja+2020, Li+2022}, while most have
        partial coverage, lacking some feature dimensions.  Most streams
        detected by \Gaia{} \citep{Gaia2016, Gaia2023}, for instance, have proper
        motions but not radial velocities nor high signal-to-noise parallax
        measurements.  Follow-up surveys, like \dataarchive{S5} \citep{Li+2019},
        add many feature dimensions, however only for a select set of streams.
        Given the diversity of coverage it is important that the likelihood
        model -- \autoref{eq:membership_prob} -- account for missing data.

        The method of treatment of missing data depends on the cause for that
        data to be missing.  Broadly, there are two categories of causes:
        randomness and systematics.  Consider a dataset with all features
        measured but with a random subset of features masked.  One possible
        approach is to attempt to impute the missing features, then feed the
        full-featured data to the likelihood model. Alternatively, the
        likelihood model might use only the present features, ignoring those
        missing. If however, features are missing for systematic reasons, due to
        a selection function, then this approach will introduce systematic bias.
        The way to avoid bias where the data is incomplete due to the selection
        function is to know that selection function. Otherwise the probabilities
        will not be properly normalized.  Selection functions are often very
        challenging to characterize, so it is routine and convenient to simply
        cut the data where the effects of the selection function become
        significant.  Not including such affected data moves the contribution of
        the selection function from a prior to the evidence.  Our models compute
        likelihoods under a specific dataset, and \autoref{eq:membership_prob}
        is a ratio of the probabilities, so the removed selection functions are
        not expected to be impactful.

        We adopt the approach of modeling the data masks as randomly-distributed
        and not systematic. We do not impute missing data, instead the per-star
        likelihood is modeled with a distribution with the same dimensionality
        as the data vector.  For example, a star with a 6D astrometric
        measurement will be modeled with a full 6D distribution, while a star
        missing its radial velocity measurement will be modeled with a
        dimensionally-reduced (5D) form of the 6D distribution. Using
        dimensionally reduced distributions ensure that missing phase-space
        dimensions do not amplify nor suppress the likelihood.  Notwithstanding,
        we include a quality flag on our calculations indicating the number of
        missing features.  We leave incorporating priors on the data mask
        distribution to future work.
